Detecting ditches using machine learning on high-resolution DEMs

### 2020-07-09 LaTeX/Ditch detection in high-resolution digital elevation models.tex
Adds updates to article after read-through. Adds updates to rejoinder.

The confidence interval for the Cohen's Kappa index ranged \textbf{[0.693 , 0.786]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represented the strongest ditch predictors.

@@

There are other studies where geographical data has been used with machine learning to map ditches. \citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds. They used several different input variables where neighbouring points of a specific LiDAR point were examined. Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii. We reproduce these neighbouring area variables when building the ditch locator in our study. Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches. These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions. In a pilot study, we compared several different algorithms (Random Forests, XGBoost, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}. Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used. To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed. This mapping was used to both train the Random Forests models and to produce the ground-truth. To evaluate the performance of the ditch locator, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.

@@

\subsection{Digitising the ground truth}
A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{Appendix} \ref{lidartodem} for specifications of the process. To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. Road ditches, forest ditches, and agricultural ditches can be found in the catchment. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600. In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches. In uncertain areas, we also found it useful to view the data in 3D. However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field. We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}. The digitisation of the ditch networks showed that there were 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.

@@

From the ditch mapping, the vector layer was rasterised so it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch. However, this made for a good compromise for the average ditch.

@@

\begin{figure} [htb!]
    \centering
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch locator.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

  \item \textbf{High Pass Median Filter} \newline
  High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM. The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window. Negative values indicate depressions in the DEM, such as ditches. A typical false positive with this index are small hollows that also show negative numbers. HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.

@@

The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions. More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{h}, and \hyperref[fig:features]{j}.

In addition to the general input variables, a number of custom input variables were also developed. Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches. This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e} and \hyperref[fig:features]{k}.

From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched, see \hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}.

\newpage

Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}. 30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{g}).

\label{impoundmentstreamremoval}
The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches. Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were marked out. After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each. The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features} \hyperref[fig:features]{f} and \hyperref[fig:features]{l}).

Overall, 81 input variables were extracted from the terrain indices. We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use. Random Forest's feature importance was used to select the most valuable input variables for different variable amounts. It was found that using 53 input variables produced the best results.

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_j_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_k_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_l_lo.jpg}}}\hspace{5pt}
    \caption{Example of 11 of the 81 input variables, in addition to ditch labels used by the models for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF Gabor, \textbf{e:} HPMF ditch amplification, \textbf{f:} HPMF ditch amplification - streams removed \textbf{g:} Sky View Factor Gabor, \textbf{h:} Sky View Factor max, radius 6, \textbf{i:} Sky View Factor non-ditch amplification  \textbf{j:} Impoundment mean, radius 3, \textbf{k:} Impoundment ditch amplification, \textbf{l:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}
\clearpage

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch locator, the raster and ditch label data of Krycklan were manually divided into 21 subsections. Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch locator to test different ways of preparing the data. This allowed the ditch locator to be evaluated on unseen data to strengthen the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.

@@

Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}. Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}. The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}. There are several hyperparameters that can be tuned to achieve better results from the model. For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}. We performed a simple hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction. Evaluating a maximum of 25 input variables for each node, and using 200 trees produced the best results. Setting the class weight to \textit{balanced} also improved the performance of the classifiers.

\newpage

@@

\begin{table} [!htb]
    \tbl{The 81 input variables used when training the models.}
    {\begin{tabular}{ll} \toprule
      Variable/Algorithm\textsuperscript{a} & Circular radii\textsuperscript{b} \\ \midrule

      Sky View Factor raw & -\\
      Sky View Factor mean & 2, 3, 4, 6 \\
      Sky View Factor median &2, 4, 6 \\
      Sky View Factor standard deviation & 2, 4, 6 \\
      Sky View Factor min & 2, 4, 6 \\
      Sky View Factor max & 2, 4, 6 \\
      Sky View Factor non ditch amplification & - \\ 
      Sky View Factor conic filter - streams removed & - \\ 
      Sky View Factor Gabor & - \\
      Sky View Factor Gabor - streams removed & -\\

      Impoundment raw & - \\
      Impoundment mean & 2, 3, 4, 6 \\
      Impoundment median & 2, 4, 6 \\
      Impoundment standard deviation & 2, 4, 6 \\
      Impoundment min & 2, 4, 6 \\
      Impoundment max & 2, 4, 6 \\
      Impoundment ditch amplification & - \\
      Impoundment ditch amplification - streams removed & - \\

      HPMF raw & - \\
      HPMF mean & 2, 3, 4, 6 \\
      HPMF median & 2, 4, 6 \\
      HPMF standard deviation & 2, 4, 6 \\
      HPMF min & 2, 4, 6 \\
      HPMF max & 2, 4, 6 \\
      HPMF ditch amplification & - \\
      HPMF ditch amplification - streams removed & - \\
      HPMF Gabor & -\\
      HPMF Gabor - streams removed & -\\

      Slope raw & - \\
      Slope mean & 2, 3, 4, 6 \\
      Slope median & 2, 4, 6 \\
      Slope standard deviation & 2, 4, 6 \\
      Slope min & 2, 4, 6 \\
      Slope max & 2, 4, 6 \\
      Slope non-ditch amplification & - \\ \bottomrule
    \end{tabular}}
    \tabnote{\textsuperscript{a} Displays algorithm used to produce the variable. \newline
        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm. The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
    \label{featuretable}
\end{table}

@@

\newpage

@@

From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch locator generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.

\begin{table}[!htb]
    \tbl{Metrics for the prediction performance of our ditch locator.}{\begin{tabular}{lccc} 
        \toprule
        Metric & Total\textsuperscript{a} & Subsection\textsuperscript{b}& CI 95\%\textsuperscript{c} \\ 
        &&Average&\\ \midrule
        Accuracy     \% & 98.96 & 98.96 & [98.64 , 99.27] \\
        Recall       \% & 77.41 & 76.81 & [69.35 , 84.27] \\
        Precision    \% & 73.89 & 73.06 & [69.73 , 76.40] \\
        $\kappa$ rating & 0.751 & 0.739 & [0.693 , 0.786] \\
        AUPRC           & 0.759 & 0.752 & [0.706 , 0.798] \\ 
        \bottomrule
    \end{tabular}}
    \tabnote{
        \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
        \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
        \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.751). The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029). Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis has been confirmed by the experiments. Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).

>>>>

The confidence interval for the Cohen's Kappa index ranged \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. Input variables based on the Impoundment Index overall represented the strongest ditch predictors.

@@

There are other studies where geographical data has been used with machine learning to map ditches. \citet{roelens} used the Random Forests algorithm to detect ditches in Belgium, using LiDAR point clouds. They used several different input variables where neighbouring points of a specific LiDAR point were examined. Values of all LiDAR attributes were represented for the neighbouring area for both 0.5, 1, 1.5, 2, 3 and 4 metre radii. We reproduce these neighbouring area variables when building the ditch detector in our study. Several other input variables for the model are calculated from the indices to place emphasis on the clear linear directional characteristics of ditches. These characteristics are also used when processing the model's output, in order to remove noise and other inconsistencies in the predictions. In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. Random Forests also computes Mean Decrease in Impurity, or Gini importance, highlighting what input variables are the most important for a given prediction \citep{gini}. Because of these factors, and the fact that it proved to be suitable for \citet{roelens}, Random Forests models are used to locate the ditches.

We hypothesise that by combining the information from all the digital terrain indices using a machine learner (Random Forests), we can improve the detection of the ditches, compared to if a single index would be used. To test this, we conduct a study using a high resolution DEM of the Krycklan Catchment in Sweden \citep{krycklancatchment}, where a high precision manual mapping has been performed. This mapping was used to both train the Random Forests models and to produce the ground-truth. To evaluate the performance of the ditch detector, we capture the various aspects of detection performance through the following measures: accuracy, precision, recall, Cohen’s Kappa index, and Area under Precision-Recall curve score.

@@

\subsection{Digitising the ground truth}
A high resolution DEM was generated from LiDAR data, see \hyperref[lidartodem]{'Data and codes availability statement'} for specifications of the process. To train and evaluate the Random Forests models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. Road ditches, forest ditches, and agricultural ditches can be found in the catchment. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. To ensure that the digitised ditch was located correctly inside the ditch and not on the bank nearby, the digitisation was conducted by zooming in to a scale of 1:600. In addition to this, both current and historical records and orthophotos dating back to 1960’s were used to help locate all the ditches. In uncertain areas, we also found it useful to view the data in 3D. However, the digitising was conducted in 2D, and in uncertain areas the ditches were verified in the field. We are confident that we have captured all the ditches in the catchment according to the digitised  ditch layers in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. Though, the length of the stream network vary substantially between season, ranging from 140 km during baseflow to 630 km during high-flow conditions \citep{mappingtemporal}. The digitisation of the ditch networks showed that there are 107 km road ditches and 208 km forest/agricultural ditches (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which aims to show median flow conditions, was 197 km.

@@

From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Because the observed average width of ditches is larger than 0.5 metres, all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the Random Forests models. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch. However, this made for a good compromise for the average ditch.

@@

\begin{figure} [htb!]
    \centering
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch detector.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

@@

  \item \textbf{High Pass Median Filter} \newline
  High Pass Median Filters (HPMF) highlight ditches by indicating local depressions in the DEM. The algorithm operates essentially by subtracting the value at the grid cell at the centre of the window from the median value in the window. Negative values indicate depressions in the DEM, such as ditches. Typical false positives with this index are small hollows that also show negative numbers. HPMF was calculated using Whitebox Tools \citep{whiteboxtools} with a window size of $4.5 * 4.5$ m.

@@

The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions. More diverse input variables were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).

In addition to the general input variables, a number of custom input variables were also developed. Several of these variables were calculated by looking at the separate digital terrain indices, attempting to amplify the ditches. This was done in several different ways for the indices by combining statistical aggregations, and by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).

From the Slope and Sky View Factor indices, input variables for the models were also extracted to attempt to amplify the values of pixels that lay in streams or in hilly areas, but leave pixels in ditches or other geographical areas untouched.

Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}. 30 Gabor filters, which were rotated in different angles and with different frequencies, were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).

\label{impoundmentstreamremoval}
The raw Impoundment Index dam height was used to create a mask, attempting to mark out streams, but retain ditches. Only areas with a relatively high dam height, above 14 metres, which would indicate that these areas most likely contained streams, were marked out. After widening the resulting area, this mask was used to attempt to remove streams from all the aforementioned custom input variables, generating one new input variable from each. The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).

Overall, 81 input variables were extracted from the terrain indices. We then conducted a sub-experiment to find the best input variables, as well as the optimal number of variables to use. Random Forest's feature importance was used to select the most valuable input variables for different variable amounts. It was found that using 40 input variables produced the best results.

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_a_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_b_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_c_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_d_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_e_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_f_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_g_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_h_lo.jpg}}}
    \subfigure[]{
        \resizebox*{4cm}{!}{\includegraphics{./images/feature_i_lo.jpg}}}
    \caption{Example of 8 of the 40 input variables, in addition to ditch labels used by the models for a small sample area. The radii represents pixels with a 0.5 m resolution. \newline \textbf{a:} Labelled ditches, \textbf{b:} Slope standard deviation, radius 6, \textbf{c:} HPMF mean, radius 4, \textbf{d:} HPMF ditch amplification, \textbf{e:} Sky View Factor Gabor, \textbf{f:} Sky View Factor max, radius 6, \textbf{g:} Impoundment mean, radius 3, \textbf{h:} Impoundment ditch amplification, \textbf{i:} Impoundment ditch amplification - streams removed}
    \label{fig:features}
\end{figure}

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections. Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data. This allowed the ditch detector to be evaluated on unseen data to strengthen the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively.

@@

Random Forests uses two techniques called \textit{subspace sampling} and \textit{bootstrap aggregation} (bagging) in tandem \citep{breiman}. Subspace sampling is the process of drawing a random subset of variables from the input variable set \citet{ho}. The bagging technique takes, with replacement, a set of different random samples for constructing each tree \citep{flach}. There are several hyperparameters that can be tuned to achieve better results from the model. For example, you can adjust the amount of decision trees to use, or the maximum amount of input variables to use for each subspace sampling \citep{scikit-learn}. We performed a hyperparameter tuning to determine what parameter values for the Random Forests algorithm would yield the best prediction. Using 300 trees and \textit{gini} as the splitting criterion, with a minimum of 10 samples per node split, and no artificial class weight or max depth of trees produced the best results.

@@

\begin{table} [!htb]
    \tbl{The 40 input variables used when training the models.}
    {\begin{tabular}{ll} \toprule
      Variable/Algorithm\textsuperscript{a} & Circular radii\textsuperscript{b} \\ \midrule

      Sky View Factor median &2, 6 \\
      Sky View Factor standard deviation & 6 \\
      Sky View Factor min & 6 \\
      Sky View Factor max & 2, 4, 6 \\
      Sky View Factor non ditch amplification & - \\ 
      Sky View Factor Gabor & - \\
      Sky View Factor Gabor - streams removed & -\\

      Impoundment mean & 2, 3, 4, 6 \\
      Impoundment median & 2, 4, 6 \\
      Impoundment standard deviation & 4, 6 \\
      Impoundment max & 6 \\
      Impoundment ditch amplification & - \\
      Impoundment ditch amplification - streams removed & - \\

      HPMF mean & 3, 4, 6 \\
      HPMF median & 4 \\
      HPMF standard deviation & 6 \\
      HPMF min & 2, 4 \\
      HPMF ditch amplification & - \\
      HPMF ditch amplification - streams removed & - \\
      HPMF Gabor - streams removed & -\\

      Slope mean & 6 \\
      Slope median & 6 \\
      Slope standard deviation & 4, 6 \\
      Slope min & 2, 4, 6 \\
      %Slope max & 2, 4, 6 \\
      Slope non-ditch amplification & - \\ \bottomrule
    \end{tabular}}
    \tabnote{\textsuperscript{a} Displays algorithm used to produce the variable. \newline
        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm. The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
    \label{featuretable}
\end{table}

@@

@@

From \hyperref[predictionperformance]{Table} \ref{predictionperformance}, it can be seen that averaging the results from the 11 subsections will yield a very similar value to the total results, indicating that the ditch detector generally performs equally well on subsections with a small amount of ditches as on subsections with a large amount of ditches.

\begin{table}[!htb]
    \tbl{Metrics for the prediction performance of our ditch detector.}{\begin{tabular}{lccc} 
        \toprule
        Metric & Total\textsuperscript{a} & Subsection\textsuperscript{b}& CI 95\%\textsuperscript{c} \\ 
        &&Average&\\ \midrule
        Accuracy     \% & 99.00 & 99.00 & [98.69 , 99.32] \\
        Recall       \% & 70.28 & 70.19 & [61.28 , 79.09] \\
        Precision    \% & 77.38 & 75.79 & [71.94 , 79.64] \\
        $\kappa$ rating & 0.732 & 0.718 & [0.655 , 0.781] \\
        AUPRC           & 0.741 & 0.733 & [0.674 , 0.791] \\ 
        \bottomrule
    \end{tabular}}
    \tabnote{
        \textsuperscript{a} The result of all 11 subsection experiments when combined. \newline
        \textsuperscript{b} An average score from the 11 different subsections that the experiment was performed on. \newline
        \textsuperscript{c} Confidence intervals at 95 \% confidence level.}
    \label{predictionperformance}
\end{table}

The $\kappa$ rating for our method can be seen in the \textit{Total} column in \hyperref[predictionperformance]{Table} \ref{predictionperformance} ($\kappa$ = 0.732). The $\kappa$ ratings from the four digital terrain indices experiments can be seen in  \hyperref[recreatedpredictionperformance]{Table} \ref{recreatedpredictionperformance} ($\kappa$ = 0.048, $\kappa$ = 0.215, $\kappa$ = 0.211, $\kappa$ = 0.029). Because the $\kappa$ rating from our method outperforms all digital terrain indices, our hypothesis has been confirmed by the experiments. Most false positives lie in stream channels (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{a, b}) and most false negatives occurred either due to ditches being too shallow (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{c, d}), or due to the attempt at removing streams using the input variables as explained in \ref{impoundmentstreamremoval} removing the deepest ditches  (\hyperref[fig:resultsillustrations]{Figure} \ref{fig:resultsillustrations} \hyperref[fig:resultsillustrations]{e, f}).

\begin{figure} [!htb]
    \centering
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_1_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_2_lo.jpg}}}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_3_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_4_lo.jpg}}}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_5_lo.jpg}}}\hspace{5pt}
    \subfigure[]{
        \resizebox*{5.5cm}{!}{\includegraphics{./images/results_illustration_6_lo.jpg}}}
    \caption{The left panel shows the hillshade for a number of subsections, and the right panel shows the hillshade with the extracted ditches superimposed on top. Green marks correctly classified ditches (true positives), red marks missed ditches (false negatives), blue marks incorrectly classified ditches (false positives), and transparent areas mark correctly classified non-ditches (true negatives). \textbf{a} and \textbf{b} illustrate that the most common false positives were natural stream channels being classified as ditches. \textbf{c} and \textbf{d} illustrate that very shallow ditches that were barely visible in the DEM were not captured with our ditch detector. \textbf{e} and \textbf{f} illustrate that some of the deepest and more prominent ditches were also missed. This occurred as a result of attempting to get rid of more false positives in the form of streams by using the input variables, as explained in \ref{impoundmentstreamremoval}.}
    \label{fig:resultsillustrations}
\end{figure}
\clearpage
